---
title: "ElMaven"
author: "Brian Yandell"
date: "2023-10-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.width = 7,
                      fit.heigth = 9)
```

```{r}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(purrr)
library(stringr)
```

# El Maven One Mouse over GTT Time

How do we find indicator of quality.

<https://elmaven.readthedocs.io/en/documentation-website/IntroductiontoElMAVENUI.html#peak-table>

## One peak only

```{r}
dir("data/data1")
```

```{r message = FALSE}
elmaven1 <- dir("data/data1") %>%
  set_names() %>%
  map(function(x) {
    out <- read_csv(file.path("data/data1", x))
    runs <- grep("run", names(out))
    if(length(runs) == 1) {
      out$label <- last(names(out))
      out$AUC <- out[[runs]]
      out$combined <- FALSE
    } else {
      out$label <- last(names(out))
      out$AUC <- NA
      out$combined <- TRUE
    }
    out[-runs]  
    }) %>%
  bind_rows() %>%
  mutate(
    minute = 
      as.numeric(str_replace(label, "run.*_([0-9]+)min.*", "\\1")),
    posneg = ifelse(grepl("_HC_", label), "HC", "RP"),
    mouse = str_remove(str_remove(label, "_20.*$"), "^.*_"),
    label = ifelse(combined, "combined", label)) %>%
  select(-combined)
```

```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(ppmDiff)) +
  aes(AUC, expectedRtDiff, col = ppmDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_sqrt() +
  scale_x_log10() +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(ppmDiff)) +
  aes(AUC, medRt + expectedRtDiff, col = ppmDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_log10() +
  scale_x_log10() +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(expectedRtDiff)) +
  aes(AUC, ppmDiff, col = expectedRtDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10() +
  facet_grid(posneg~minute)  +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(expectedRtDiff)) +
  aes(AUC, medMz+ ppmDiff, col = expectedRtDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_log10()  +
  scale_x_log10() +
  facet_grid(posneg~minute)  +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(desc(maxQuality))) +
  aes(AUC, medMz+ ppmDiff, col = maxQuality)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_log10()  +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(medRt)) +
  aes(AUC, maxQuality, col = medRt)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven1 %>%
         filter(label != "combined") %>%
         arrange(desc(AUC))) +
  aes(medRt, maxQuality, col = AUC)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

\newpage

### Combined vs Minute Measurements

Six time points run together.
Note that `parent` and `medMz` are identical in each dataset, but `medMz` and `medRt`
vary for a compound across datasets.
There may be two entries for some compounds.
First look at those and then reduce to compounds with only one entry.

```{r}
elmaven1 %>%
  filter(label == "combined") %>%
  count(compound, posneg) %>%
  count(posneg, n, name = "count")
```

```{r}
elmaven1 %>%
  filter(label != "combined") %>%
  count(compound, posneg, minute) %>%
  count(posneg, minute, n, name = "count") %>%
  pivot_wider(names_from = "minute", values_from = "count")
```

```{r}
elmaven1 %>%
  filter(label == "combined") %>%
  count(compound, posneg) %>%
  filter(n > 1)
```

```{r}
elmaven1 %>%
  filter(label != "combined") %>%
  count(compound, posneg, minute) %>%
  filter(n > 1) %>%
  pivot_wider(names_from = "minute", values_from = "n")
```

```{r}
comb_dup <- (elmaven1 %>%
  filter(label == "combined") %>%
  count(compound, posneg) %>%
  filter(n > 1))$compound
min_dup <- (elmaven1 %>%
  filter(label != "combined") %>%
  count(compound, posneg, minute) %>%
  filter(n > 1) %>%
  pivot_wider(names_from = "minute", values_from = "n"))$compound
all_dup <- unique(c(comb_dup, min_dup))
```

\newpage

```{r}
combined <- elmaven1 %>%
  filter(label == "combined", !(compound %in% all_dup)) %>%
  select(compound, posneg,
         medMz, medRt,
         maxQuality, expectedRtDiff, ppmDiff)
```

```{r}
elmaven1_comb <- left_join(
  elmaven1 %>%
    filter(label != "combined", !(compound %in% all_dup)) %>%
    select(compound, posneg, minute,
           medMz, medRt,
           maxQuality, expectedRtDiff, ppmDiff),
  combined,
  by = c("compound", "posneg"),
  suffix = c("_min", "_comb"),
  relationship = "many-to-one"
)
```

```{r warning = FALSE}
ggplot(elmaven1_comb %>% arrange(desc(maxQuality_comb))) +
  aes(medRt_comb, medRt_min - medRt_comb, col = maxQuality_comb) +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven1_comb %>% arrange(desc(maxQuality_comb))) +
  aes(medMz_comb, medMz_min - medMz_comb, col = maxQuality_comb) +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven1_comb) +
  aes(maxQuality_comb, maxQuality_min - maxQuality_comb) +
  geom_point(size = 2, alpha = 0.25) +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

\newpage

## Two peaks possible

```{r}
dir("data/data2")
```

```{r message = FALSE}
elmaven <- dir("data/data2") %>%
  set_names() %>%
  map(function(x) {
    out <- read_csv(file.path("data/data2", x))
    runs <- grep("run", names(out))
    if(length(runs) == 1) {
      out$label <- last(names(out))
      out$AUC <- out[[runs]]
      out$combined <- FALSE
    } else {
      out$label <- last(names(out))
      out$AUC <- NA
      out$combined <- TRUE
    }
    out[-runs]  
    }) %>%
  bind_rows() %>%
  mutate(
    minute = 
      as.numeric(str_replace(label, "run.*_([0-9]+)min.*", "\\1")),
    posneg = ifelse(grepl("_HC_", label), "HC", "RP"),
    mouse = str_remove(str_remove(label, "_20.*$"), "^.*_"),
    label = ifelse(combined, "combined", label)) %>%
  select(-combined)
```

```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(ppmDiff)) +
  aes(AUC, expectedRtDiff, col = ppmDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_sqrt() +
  scale_x_log10() +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(ppmDiff)) +
  aes(AUC, medRt + expectedRtDiff, col = ppmDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_log10() +
  scale_x_log10() +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(expectedRtDiff)) +
  aes(AUC, ppmDiff, col = expectedRtDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10() +
  facet_grid(posneg~minute)  +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(expectedRtDiff)) +
  aes(AUC, medMz+ ppmDiff, col = expectedRtDiff)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_log10()  +
  scale_x_log10() +
  facet_grid(posneg~minute)  +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(desc(maxQuality))) +
  aes(AUC, medMz+ ppmDiff, col = maxQuality)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_y_log10()  +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(medRt)) +
  aes(AUC, maxQuality, col = medRt)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(desc(AUC))) +
  aes(medRt, maxQuality, col = AUC)+
  geom_smooth(se = FALSE, method = "lm", col = "red", size = 2,
              formula = "y ~ x") +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

\newpage

### Combined vs Minute Measurements

Six time points run together.
Note that `parent` and `medMz` are identical in each dataset, but `medMz` and `medRt`
vary for a compound across datasets.
There may be two entries for some compounds.
First look at those and then reduce to compounds with only one entry.

```{r}
elmaven %>%
  filter(label == "combined") %>%
  count(compound, posneg) %>%
  count(posneg, n, name = "count")
```

```{r}
elmaven %>%
  filter(label != "combined") %>%
  count(compound, posneg, minute) %>%
  count(posneg, minute, n, name = "count") %>%
  pivot_wider(names_from = "minute", values_from = "count")
```

```{r}
elmaven %>%
  filter(label == "combined") %>%
  count(compound, posneg) %>%
  filter(n > 1)
```

```{r}
elmaven %>%
  filter(label != "combined") %>%
  count(compound, posneg, minute) %>%
  filter(n > 1) %>%
  pivot_wider(names_from = "minute", values_from = "n")
```

```{r}
comb_dup <- (elmaven %>%
  filter(label == "combined") %>%
  count(compound, posneg) %>%
  filter(n > 1))$compound
min_dup <- (elmaven %>%
  filter(label != "combined") %>%
  count(compound, posneg, minute) %>%
  filter(n > 1) %>%
  pivot_wider(names_from = "minute", values_from = "n"))$compound
all_dup <- unique(c(comb_dup, min_dup))
```

\newpage

```{r}
combined <- elmaven %>%
  filter(label == "combined", !(compound %in% all_dup)) %>%
  select(compound, posneg,
         medMz, medRt,
         maxQuality, expectedRtDiff, ppmDiff)
```

```{r}
elmaven_comb <- left_join(
  elmaven %>%
    filter(label != "combined", !(compound %in% all_dup)) %>%
    select(compound, posneg, minute,
           medMz, medRt,
           maxQuality, expectedRtDiff, ppmDiff),
  combined,
  by = c("compound", "posneg"),
  suffix = c("_min", "_comb"),
  relationship = "many-to-one"
)
```

```{r warning = FALSE}
ggplot(elmaven_comb %>% arrange(desc(maxQuality_comb))) +
  aes(medRt_comb, medRt_min - medRt_comb, col = maxQuality_comb) +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven_comb %>% arrange(desc(maxQuality_comb))) +
  aes(medMz_comb, medMz_min - medMz_comb, col = maxQuality_comb) +
  geom_point(size = 2, alpha = 0.25) +
  scale_x_log10()  +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

```{r warning = FALSE}
ggplot(elmaven_comb) +
  aes(maxQuality_comb, maxQuality_min - maxQuality_comb) +
  geom_point(size = 2, alpha = 0.25) +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

\newpage

```{r}
knitr::knit_exit()
```

Interactive version

```{r}
plotly::ggplotly(
ggplot(elmaven %>%
         filter(label != "combined") %>%
         arrange(ppmDiff)) +
  aes(AUC, expectedRtDiff, col = ppmDiff, compound = compound)+
  geom_point(size = 2, alpha = 0.25) +
  scale_y_sqrt() +
  scale_x_log10() +
  facet_grid(posneg~minute) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
)
```

