---
title: "QC_Automation"
author: "Brian Yandell"
date: "2024-09-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(metabr)
```

# QC steps and normalization (Batch Correction)

Need to correct the two modes (`HC_Neg` and `RP_Pos`) separately.
`HC_Neg` mode targeted data of founder samples used as example
to showed how Qiushi did batch correction manually.
See Batch_Correction_Procedure_RC2_Project.pdf
(file `data/QC Batch Correction/Batch_Correction_Procedure_RC2_Project.pdf`)

```{r}
# Raw data in CSV
dirpath <- 'data/QC Batch Correction/#Founder_Plasma_Raw_Data'
# Raw and Corrected data in XLSX
handpath <- 'data/QC Batch Correction/#Founder_Plasma_Batch_Corrected_Data'
```

## Read Hand-calculated Targeted Data

These are as CSV files and as sheets of XLSX files.

```{r warning=FALSE}
hand_HCneg <- read_batch_metab(handpath,
  "Founder_Plasma_HC_Neg_Targeted_Batch_Corrected_Data.xlsx")
```

```{r warning=FALSE}
hand_RPpos <- read_batch_metab(handpath,
  "Founder_Plasma_RP_Pos_Targeted_Batch_Corrected_Data.xlsx")
```

## Individual QC Steps

Shown for `HC_Neg` data.
For `RP_Pos` data, exclude every 1st QC run on plate as it
historically was always an outlier.
QC steps were verified against hand-calculations.
See [QC_steps.Rmd](https://github.com/byandell/metabr/blob/master/QC_steps.Rmd)
section titles `Examine QC data`.

1. Transpose Data and Extract Info from Sample Names

Note addition of `BatchPlate_HCneg` as QC is by `Plate` and `Batch`, not `rundate`.

```{r}
raw_HCneg <- read_raw_metab(dirpath,
  "Founder_Plasma_All_Timepoints_HC_Neg_targeted_Raw.csv",
  Batch_Plate = BatchPlate_HCneg)
```

2. Only pick up QC runs of every plate, and calculate correction factors (CF) for each metabolite.
    + Average of QC per plate/Average of QC all plate.

```{r}
qc_HCneg <- calc_cf(raw_HCneg)
```

3. If metabolites weren’t measured in (**all?**) QC samples, but were measured in study samples: use the CF average of all metabolites in one plate as the CF of the missing metabolites on that plate.

```{r}
qc_HCneg <- replace_missing_ave_cf(qc_HCneg)
```

4. Correct the Peak Area using CF for each metabolite 

```{r}
rescale_HCneg <- correct_raw_cf(raw_HCneg, qc_HCneg)
```

## Combine QC Steps Into One Routine.

```{r}
rescale_HCneg <- qc_steps(dirpath,
  "Founder_Plasma_HC_Neg_Targeted_Batch_Corrected_Data.xlsx",
  FALSE, BatchPlate_HCneg)
```

```{r}
rescale_RPpos <- qc_steps(dirpath,
                          "Founder_Plasma_All_Timepoints_RP_Pos_targeted_Raw.csv",
                          exclude_first = TRUE)
```

# Untargeted Data

Still to do.

