---
title: "Metabolite QC Analysis"
author: "Brian Yandell"
date: "2023-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```
From: Sun, Qiushi <qiushi.sun@yale.edu>
Date: Monday, November 27, 2023 at 2:02 PM
To: Brian Yandell <brian.yandell@wisc.edu>, Kibbey, Richard <richard.kibbey@yale.edu>, Cardone, Rebecca <rebecca.cardone@yale.edu>
Subject: Re: Checking in after my visit

I had finished writing down the step-by-step instructions for the following two bottlenecks:

- flux data from mass spec to elmaven on Yale cluster
- QC steps and normalization (QC Batch Correction)

The examples of raw data and batch corrected data were also ready to share. However, I had issues with our metabolomics notebook to normalize and merge targeted and untargeted data, which Mano is working on fixing it. She said she would have it done this week. I can share all other files I've prepared with you today. And once I get the normalized data, I will update you.
 
About "quality score vs deviation of Rt from theory", Elucidata is still working on developing a notebook to do the calculation. I should get more information on this after our Wednesday meeting with Elucidata. 
 
In addition, Elucidata needs more clarification on how we want to visualize metabolites and enzymes in pathways. I think it's better to discuss this in Wed bioinformatics meeting when you, Mark and Alan are all present. 

From: Brian Yandell <brian.yandell@wisc.edu>
Sent: Monday, November 27, 2023 01:20 PM
To: Kibbey, Richard <richard.kibbey@yale.edu>; Cardone, Rebecca <rebecca.cardone@yale.edu>; Sun, Qiushi <qiushi.sun@yale.edu>
Subject: Checking in after my visit
I am just checking in after my visit. Hope you are all well. I had a good visit to Bar Harbor, but I won’t bore you with the details here.
 
I believe we identified mass spec pipeline bottlenecks related to Qiushi’s excellent work. I wonder how we might move forward on these. Here are my notes:

- 3 bottlenecks
  + flux data from mass spec to elmaven on Yale cluster
  + QC steps and normalization
  + peak identification (mz, Rt, AUC)
- prioritization of peaks to recheck
  + quality score vs deviation of Rt from theory, etc.
  + run through then post check based on interest, significance
  + check by plate, GTT minute, other grouping

It would help me if you could write this up into a short document (1-2 pages) so that we can decide how to move forward. Can we plan to discuss next steps with these next Thu 7 Dec when we have our UW-Yale checkin?
```

# flux data from mass spec to elmaven on Yale cluster

This is a step-by-step process that is time consuming but not complicated. It relies on a few other tools. Need a (drag-and-drop?) app to automate.

# QC steps and normalization (Batch Correction)

Need to correct the two modes (`HC_Neg` and `RP_Pos`) separately.
`HC_Neg` mode targeted data of founder samples used as example
to showed how Qiushi did batch correction manually.

## Batch Correction on Targeted Data

1. Transpose Data and Extract Info from Sample Names
2. Only pick up QC runs of every plate, and calculate correction factors (CF) for each metabolite.
    + Average of QC per plate/Average of QC all plate.
3. If metabolites weren’t measured in QC samples, but were measured in study samples: use the CF average of all metabolites in one plate as the CF of the missing metabolites on that plate.
4. Correct the Peak Area using CF for each metabolite 

### Notes for batch correction in `RP_Pos` mode 

- Exclude every 1st QC run on plate for CF calculation of `RP_Pos` data as it historically was always an outlier.
- Some metabolites were not measured in QC samples and need to use average CF of all the other metabolites as their CF per plate:
  + ACETOACETATE, URACIL, SUCCINATE, PURINE, HYPOXANTHINE, 4-ACETAMIDOBUTANOATE, AMINOHYDROXYBENZOATE, DOPAMINE, OROTATE, SALSOLINOL, 3,4 DIHYDROXYMANDELATE, 3-METHOXYTYROSINE, DEOXYURIDINE, DEOXYADENOSINE, INOSINE, GUANOSINE, DEOXYADENOSINE MONOPHOSPHATE, THYROTROPIN RELEASING HORMONE, S-ADENOSYLHOMOCYSTEINE (1)   
- For untargeted data, if any metabolite has peak area <1000 or correction factor is 0 in most QC samples, use average of CF of all metabolites as their CF no matter if they can be measured in QC samples or not.

### Extra Notes for Founder Samples Runs

- There were two runs for `B6-1_120min` time point in both modes. Should use average of the two injections as the final data for this sample for both targeted and untargeted.
  + `run11-27Jul21_B6-1_20210106_120min_1_HC_Neg/RP_Pos`
  + `run108-27Jul21_B6-1_20210106_120min_2_HC_Neg/RP_Pos`
- There was an accident interruption when running `batch 1 plate 7 in HC_Neg` mode; this plate had two run dates:
  + 6-Aug-21
  + 7-Aug-21

## Batch Correction on Untargeted Data

### Annotation

For untargeted curation, we usually assign up to the five most intense peaks for each metabolite in each mode based on their `m/z`. One metabolite may have multiple entries which can be distinguished by the different retention times. We need to do batch correction on each entry separately. 

### QC Steps for Untargeted Data

1. Transpose Data and Extract Info from Sample Names, should use `compound name + retention time` as the compound header to distinguish different entries of the same metabolite.
2. Same as targeted, only pick QC runs in each plate and calculate CF  
3. Correct the Peak Area using corresponding CF for each metabolite 

### Data Normalization and Merge

Annotation

- After batch correction, need to run a notebook on `Elucidata Polly` to normalize and merge targeted and untargeted data.
- Data normalization: After this step, all values are negative. Do normalization in each mode, targeted and untargeted separately. 
  + normalize each metabolite by the sum of all metabolites in a sample per mode
  + log2 transformation
- Merge targeted and untargeted data to remove replicates:
  + For metabolites which both targeted and untargeted curation detected, only retain the targeted data since it’s more reliable. 
  + For metabolites which have multiple entries, only retain the largest normalized data.
- The final data will be only one entry for each metabolite
  + all-capitalized metabolites are from targeted curation
  + non-capitalized metabolites are from untargeted curation

1. Need to transpose corrected data into the format that notebook requires (the same format as original output file from `El-maven`)
    + Can’t add or delete columns since the notebook recognizes data by column numbers. 
    + Delete all QC runs.
2. Things to do before running notebook:  
    + Calculate average of the two samples `run11-27Jul21_B6-1_20210106_120min_1_HC_Neg/RP_Pos` and `run108-27Jul21_B6-1_20210106_120min_2_HC_Neg/RP_Pos` as the final data for this time point in both targeted and untargeted data.



# QC Analysis with R Script

```{r}
dirpath <- 'data/QC Batch Correction/#Founder_Plasma_Raw_Data/'
```

## Targeted HC_Neg

1. Transpose Data and Extract Info from Sample Names

```{r}
HCneg_t <- 
  # Data are in sheet 1 starting on line 1.
  readr::read_csv(file.path(dirpath, "Founder_Plasma_All_Timepoints_HC_Neg_targeted_Raw.csv")) |>
    
  # Pivot traits, which begin after `parent` column.
  tidyr::pivot_longer(-(label:parent), names_to = "mouse_id", values_to = "value") |>
  dplyr::mutate(run = stringr::str_replace(mouse_id, "^run([0-9]+)-.*", "\\1"),
                rundate = stringr::str_replace(mouse_id, "^run[0-9]+-([0-9A-Za-z]+)_.*", "\\1")) 
```

2. Only pick up QC runs of every plate, and calculate correction factors (CF) for each metabolite.
    + Average of QC per plate/Average of QC all plate.

```{r}
HCneg_t_qc <- HCneg_t |>
  dplyr::filter(stringr::str_detect(mouse_id, "_QC_")) |>
  dplyr::group_by(rundate, compound) |>
  dplyr::summarize(value = mean(value, na.rm = TRUE), .groups = "drop") |>
  dplyr::ungroup()
```

```{r}
HCneg_t_qc <- 
  # Join data with `compound` averages over `rundates`.
  dplyr::left_join(
    HCneg_t_qc,
    
    # average across `rundate` by compound.
    HCneg_t_qc |>
      dplyr::group_by(compound) |>
      dplyr::summarize(value = mean(value, na.rm = TRUE), .groups = "drop") |>
      dplyr::ungroup(),
  by = "compound", suffix = c("", "_ave")) |>
  
  # Divide `value` by `value_ave` (or make NA if `value_ave` is missing).
  dplyr::mutate(value = ifelse(value_ave != 0 & !is.na(value_ave),
                               value / value_ave, NA)) |>
  dplyr::select(-value_ave)
```

3. If metabolites weren’t measured in QC samples, but were measured in study samples: use the CF average of all metabolites in one plate as the CF of the missing metabolites on that plate.

```{r}
HCneg_t_qc_date <- HCneg_t_qc |>
  # Filter out `NA` and `0` values.
  dplyr::filter(!is.na(value), value != 0) |>
  dplyr::group_by(rundate) |>
  dplyr::summarize(value = mean(value), .groups = "drop") |>
  dplyr::ungroup()
```

```{r}
# Bads are compounds with missing `value` or `value` = 0.
(bads <- HCneg_t_qc |>
  dplyr::filter(is.na(value) | value == 0) |>
  dplyr::count(compound))
```

```{r}
HCneg_t_qc <- 
  # Join `rundata` averages to QC.
  dplyr::left_join(HCneg_t_qc, HCneg_t_qc_date, by = "rundate", suffix = c("", "_ave")) |>
  # If bad `compound`, replace `value` by `value_ave`
  dplyr::mutate(value = ifelse(compound %in% bads$compound, value_ave, value)) |>
  dplyr::select(-value_ave)
```

4. Correct the Peak Area using CF for each metabolite 

```{r}
HCneg_t <-
  dplyr::left_join(HCneg_t, HCneg_t_qc, by = c("rundate", "compound"), suffix = c("", "_cf")) |>
  # Divide `value` by `value_qc` (or make NA if `value_ave` is missing, which should not happen).
  dplyr::mutate(value = ifelse(value_cf != 0 & !is.na(value_cf),
                               value / value_cf, NA)) |>
  dplyr::select(-value_cf)
```

## Targeted RP_Pos

Need to be clear what it means to be "first run QC".
