---
title: "Metabolite QC Analysis"
author: "Brian Yandell"
date: "2023-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(metabr)
```

# QC steps and normalization (Batch Correction)

Need to correct the two modes (`HC_Neg` and `RP_Pos`) separately.
`HC_Neg` mode targeted data of founder samples used as example
to showed how Qiushi did batch correction manually.

```{r}
dirpath <- 'data/QC Batch Correction/#Founder_Plasma_Raw_Data'
```

```{r}
handpath <- 'data/QC Batch Correction/#Founder_Plasma_Batch_Corrected_Data'
```

## Check Hand-calculated Targeted Data

```{r warning=FALSE}
hand_HCneg <- read_batch_metab(handpath,
  "Founder_Plasma_HC_Neg_Targeted_Batch_Corrected_Data.xlsx")
```

```{r warning=FALSE}
hand_RPpos <- read_batch_metab(handpath,
  "Founder_Plasma_RP_Pos_Targeted_Batch_Corrected_Data.xlsx")
```

Check out Batch3 entries.

```{r}
hand_HCneg |> dplyr::select(mouse_id,sample,rundate,Batch,Plate) |>
  dplyr::filter(grepl("Batch", sample)) |>
  dplyr::mutate(batch3 = stringr::str_remove(
        stringr::str_remove(.data$sample, "^.*_Batch"), "_.*$")) |>
  dplyr::count(rundate,Batch,batch3, Plate)
```

```{r}
hand_HCneg |> dplyr::select(mouse_id,sample,rundate,Batch,Plate) |>
  dplyr::filter(Batch == 2, Plate %in% c("4","7")) |>
  dplyr::count(rundate,Batch, Plate)
```

```{r}
hand_HCneg |> dplyr::select(mouse_id,sample,rundate,Batch,Plate) |>
  dplyr::filter(!(Batch == 2 & Plate %in% c("4","7"))) |>
  dplyr::count(rundate,Batch, Plate)
```

```{r}
(runBatchPlate <- hand_HCneg |>
   dplyr::distinct(rundate, Batch, Plate) |>
   tidyr::unite(Batch_Plate, Batch, Plate, remove = FALSE))
```

## Batch Correction on HC_Neg Targeted Data

1. Transpose Data and Extract Info from Sample Names

Note addition of `runBatchPlate` as QC is by `Plate` and `Batch`, not `rundate`.

```{r}
raw_HCneg <- read_raw_metab(dirpath,
  "Founder_Plasma_All_Timepoints_HC_Neg_targeted_Raw.csv",
  Batch_Plate = runBatchPlate)
```

2. Only pick up QC runs of every plate, and calculate correction factors (CF) for each metabolite.
    + Average of QC per plate/Average of QC all plate.

```{r}
qc_HCneg <- calc_cf(raw_HCneg)
```

3. If metabolites weren’t measured in (**all?**) QC samples, but were measured in study samples: use the CF average of all metabolites in one plate as the CF of the missing metabolites on that plate.

```{r}
qc_HCneg <- replace_missing_ave_cf(qc_HCneg)
```

4. Correct the Peak Area using CF for each metabolite 

```{r}
rescale_HCneg <- correct_raw_cf(raw_HCneg, qc_HCneg)
```

Now redo as one step:

```{r}
rescale_HCneg <- qc_steps(dirpath,
  "Founder_Plasma_All_Timepoints_HC_Neg_targeted_Raw.csv",
  FALSE, runBatchPlate)
```

Verify that CSV and first sheet of XLS are the same raw data.

```{r}
rescale_HCnegx <- qc_steps(handpath,
  "Founder_Plasma_HC_Neg_Targeted_Batch_Corrected_Data.xlsx",
  FALSE, runBatchPlate)
```

```{r}
all.equal(rescale_HCneg, rescale_HCnegx)
```

```{r}
rm(rescale_HCnegx)
```

## Batch Correction on RP_Pos Targeted Data

Exclude every 1st QC run on plate for CF calculation of `RP_Pos` data as it historically was always an outlier. *exclude run4 on every single plate; use average of others*

```{r}
rescale_RPpos <- qc_steps(dirpath,
                          "Founder_Plasma_All_Timepoints_RP_Pos_targeted_Raw.csv",
                          exclude_first = TRUE)
```

```{r}
rescale_HCneg |> dplyr::select(mouse_id,sample,rundate) |>
  dplyr::filter(grepl("Batch", sample)) |>
  dplyr::mutate(batch3 = stringr::str_remove(
        stringr::str_remove(.data$sample, "^.*_Batch"), "_.*$")) |>
  dplyr::count(rundate,batch3)
```

#### Examine QC data

2. Only pick up QC runs of every plate, and calculate correction factors (CF) for each metabolite.
    + Average of QC per plate/Average of QC all plate.

```{r}
qc_HCneg <- calc_cf(raw_HCneg)
```

3. If metabolites weren’t measured in (**all?**) QC samples, but were measured in study samples: use the CF average of all metabolites in one plate as the CF of the missing metabolites on that plate.

```{r}
qc_HCneg <- replace_missing_ave_cf(qc_HCneg)
```

```{r}
qc_HCneg |> dplyr::count(rundate)
```

Pull CF calculations from spreadsheet

```{r}
readxl::excel_sheets(file.path(handpath,
  "Founder_Plasma_HC_Neg_Targeted_Batch_Corrected_Data.xlsx"))
```

```{r}
qc_calc <- readxl::read_excel(file.path(handpath,
  "Founder_Plasma_HC_Neg_Targeted_Batch_Corrected_Data.xlsx"),
  sheet = 3, skip = 130, n_max = 14) |>
  dplyr::rename(Batch_Plate = "...1") |>
  dplyr::filter(!is.na(Batch_Plate)) |>
  tidyr::pivot_longer(-Batch_Plate, names_to = "compound", values_to = "value") |>
  dplyr::mutate(Batch_Plate =
    stringr::str_replace(Batch_Plate, "Batch([0-9])_Plate([0-9])", "\\1_\\2")) |>
  tidyr::separate_wider_delim(Batch_Plate, "_", names = c("Batch", "Plate"),
                              cols_remove = FALSE) |>
  dplyr::filter(!stringr::str_detect(compound, "Average"))
```

```{r}
m <- match(unique(qc_calc$Batch_Plate), unique(runBatchPlate$Batch_Plate))
```

#### Compare hand and automated Batch Correction

```{r}
out <- hand_auto(hand_HCneg, rescale_HCneg)
```

```{r}
plot(out |> dplyr::filter(abs(1 - ratio) > 1e-1,
                          !(Plate == "7" & Batch == "1"))) +
  ggplot2::ggtitle("All but Batch 1, Plate 7")
```

```{r}
plot(out |> dplyr::filter(abs(1 - ratio) > 1e-1,
                          (Plate == "7" & Batch == "1"))) +
  ggplot2::ggtitle("Only Batch 1, Plate 7")
```

```{r}
out |>
  dplyr::filter(abs(1 - ratio) > 1e-1) |>
  dplyr::count(Batch, Plate)
```

```{r}
out |>
  dplyr::filter(abs(1 - ratio) > 1e-1) |>
  dplyr::select(mouse_id, Batch, Plate, ratio)
```

```{r}
out |>
  dplyr::filter(abs(1 - ratio) > 1e-1) |>
  dplyr::count(mouse_id) |>
  dplyr::arrange(dplyr::desc(n))
```

```{r}
hand_auto(hand_RPpos, rescale_RPpos)
```

```{r}
hand_auto(hand_RPpos |> dplyr::filter(rundate == hand_RPpos$rundate[1]),
          rescale_RPpos |> dplyr::filter(rundate == hand_RPpos$rundate[1]))
```

```{r}
hand_auto(hand_HCneg |> dplyr::filter(rundate == hand_HCneg$rundate[1]),
          rescale_HCneg |> dplyr::filter(rundate == hand_HCneg$rundate[1]))
```

```{r}
hands <- hand_HCneg |> dplyr::filter(rundate == hand_HCneg$rundate[1])
autos <- rescale_HCneg |> dplyr::filter(rundate == hand_HCneg$rundate[1])
out <- dplyr::full_join(
    dplyr::select(hands, mouse_id, sample, compound, value, Batch, Plate),
    dplyr::select(autos, mouse_id, sample, compound, value),
    by = c("mouse_id", "sample", "compound"),
    suffix = c("_hand", "_auto")) |>
  dplyr::filter(value_hand > 1.1 * value_auto)
```

```{r}
hands <- hand_RPpos |> dplyr::filter(rundate == hand_RPpos$rundate[1])
autos <- rescale_RPpos |> dplyr::filter(rundate == hand_RPpos$rundate[1])
out <- dplyr::full_join(
    dplyr::select(hands, mouse_id, sample, compound, value, Batch, Plate),
    dplyr::select(autos, mouse_id, sample, compound, value),
    by = c("mouse_id", "sample", "compound"),
    suffix = c("_hand", "_auto")) |>
  dplyr::filter(value_hand > 3 * value_auto)
```

```{r}

```

### Notes for batch correction in `RP_Pos` mode 
 
- For untargeted data, if any metabolite has peak area <1000 or correction factor is 0 in most QC samples, use average of CF of all metabolites as their CF no matter if they can be measured in QC samples or not.

### **Extra Notes for Founder Samples Runs**

**This was not attempted yet.**

- There were two runs for `B6-1_120min` time point in both modes. Should use average of the two injections as the final data for this sample for both targeted and untargeted.
  + `run11-27Jul21_B6-1_20210106_120min_1_HC_Neg/RP_Pos`
  + `run108-27Jul21_B6-1_20210106_120min_2_HC_Neg/RP_Pos`
- There was an accident interruption when running `batch 1 plate 7 in HC_Neg` mode; this plate had two run dates:
  + 6-Aug-21
  + 7-Aug-21

## Batch Correction on Untargeted Data

For untargeted curation, we usually assign up to the five most intense peaks for each metabolite in each mode based on their `m/z`. One metabolite may have multiple entries which can be distinguished by the different retention times. We need to do batch correction on each entry separately. 
This is done in routine `calc_cf()` with grouping by `compound` and `medRt`.

1. Transpose Data and Extract Info from Sample Names, should use `compound name + retention time` as the compound header to distinguish different entries of the same metabolite.
2. Same as targeted, only pick QC runs in each plate and calculate CF  
3. Correct the Peak Area using corresponding CF for each metabolite 

```{r}
rescale_HCnegu <- qc_steps(dirpath, "Founder_Plasma_All_Timepoints_HC_Neg_untargeted_Raw.csv")
rescale_RPposu <- qc_steps(dirpath, "Founder_Plasma_All_Timepoints_RP_pos_untargeted_Raw.csv",
                           exclude_first = TRUE)
```

## Data Normalization and Merge

```{r}
ggplot2::ggplot(rescale_HCneg) +
  ggplot2::aes(value) +
  ggplot2::geom_density() +
  ggplot2::scale_x_log10()
```

Annotation

- After batch correction, need to run a notebook on `Elucidata Polly` to normalize and merge targeted and untargeted data.
- Data normalization: After this step, all values are negative. Do normalization in each mode, targeted and untargeted separately. 
  + normalize each metabolite by the sum of all metabolites in a sample per mode
  + log2 transformation
  
```{r}
norm_HCneg <- normalize(rescale_HCneg)
```

```{r}
ggplot2::ggplot(norm_HCneg) +
  ggplot2::aes(value) +
  ggplot2::geom_density()
```

- Merge targeted and untargeted data to remove replicates:
  + **For metabolites which both targeted and untargeted curation detected, only retain the targeted data since it’s more reliable.** 
  + **For metabolites which have multiple entries, only retain the largest normalized data.**

```{r}
all_data <- 
  dplyr::bind_rows(
    HC_Neg__yes = rescale_HCneg,
    HC_Neg__no = rescale_HCnegu,
    RP_Pos__yes = rescale_RPpos,
    RP_Pos__no = rescale_RPposu,
    .id = "mode") |>
  tidyr::separate_wider_delim(
    mode, "__", names = c("mode","targeted"))
```

** better to use KEGG ID (compoundId) instead of compound; names might be different **
** if there is no KEGG ID, use compound name **
** ultimately use compound name for further work **
** but might want dictionary **

```{r}
distinct_compounds <- all_data |>
  dplyr::group_by(.data$mode) |>
  dplyr::distinct(.data$compoundId, .data$targeted) |>
  dplyr::count(.data$compoundId, .data$targeted) |>
  tidyr::pivot_wider(
    names_from = "targeted", values_from = "n",
    values_fill = 0) |>
  dplyr::filter(.data$yes + .data$no == 2) |>
  dplyr::ungroup()
```

**Now need to remove untargeted that have these names**.

```{r}
table(distinct_compounds$mode)
```

```{r}
dplyr::distinct(all_data, mode, compoundId, medRt) |>
  dplyr::count(mode, compoundId) |>
  dplyr::count(mode, n, name = "medRt") |>
  tidyr::pivot_wider(names_from = "mode", values_from = "medRt", values_fill = 0)
```

Combine HC and RP and get "best" ones. 

1. Look for targeted; keep only in targeted
2. Same compound in HC_Neg and RP_Pos would have 2 values; take one with "best" signal.
Larger peak area (ignore retention time) in some sense.

```{r}
wh_max <- dplyr::group_by(all_data, mode, compound, medRt) |>
  dplyr::summarize(value = mean(value), .groups = "drop") |>
  dplyr::ungroup() |>
  dplyr::group_by(mode, compound) |>
  dplyr::summarize(
    medRt = .data$medRt[which.max(.data$value)][1],
    .groups = "drop") |>
  dplyr::ungroup()
```

**Now want to select entries for compound from mode with the
chosen medRt.**

```{r}
```

- The final data will be only one entry for each metabolite
  + all-capitalized metabolites are from targeted curation
  + non-capitalized metabolites are from untargeted curation

1. Need to transpose corrected data into the format that notebook requires (the same format as original output file from `El-maven`)
    + **Can’t add or delete columns since the notebook recognizes data by column numbers.**
    + (This requires some sort of left_join to get back to original order?) 
    + (**Does it require rows to be in same order, or just columns?**)
    + Delete all QC runs.
2. Things to do before running notebook:  
    + Calculate average of the two samples `run11-27Jul21_B6-1_20210106_120min_1_HC_Neg/RP_Pos` and `run108-27Jul21_B6-1_20210106_120min_2_HC_Neg/RP_Pos` as the final data for this time point in both targeted and untargeted data.

## Targeted RP_Pos

Need to be clear what it means to be "first run QC".
